A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-26 imaging_demo/ms.tex
Incorporate many of the first round of suggestions
From a5f2570 (master):
=====================

(*) Add LWA_MOFF_bandavg_image_4_iterations_analytic_aperture.eps to
repository and remove LWA_sample_image_391_iterations.png from the
repository

In epic.bib,

(*) Add Bowman et al.(2013) to references

In ms.tex,

(*) Many many of the changes suggested by Judd and Adam
(*) Modify table

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are marching towards densely packed array layouts 
consisting of a large number of antennas ($N$). Since the cost of traditional 
correlators scales as $\mathcal{O}(N^2)$, there will be a steep cost for 
realizing the full imaging potential of these powerful instruments. Through 
our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present 
the first software demonstration of a generalized direct imaging algorithm 
known as the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost to $\mathcal{O}(N\log N)$ but can also image 
from irregularly arranged heterogeneous antenna arrays. EPIC is highly modular and 
parallelizable. It is implemented in object oriented Python and is publicly 
available. We have verified the images produced to be mathematically identical 
to those produced using traditional techniques. We have also validated our 
implementation on data observed with the Long Wavelength Array. Antenna systems
with a dense filling factor consisting of a large number of antennas such as 
LWA, SKA, and HERA will gain a significant advantage by deploying EPIC. 
Inherent availability of calibrated time-domain images on timescales roughly 
equal to the writeout timescale of the digitizer will make it a prime 
candidate for transient searches of primarily Fast Radio Bursts (FRB) as well 
as slower planteary and exoplantenary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction. A number of new telescopes are under development around the world
based on this new paradigm, including the Murchison Widefield Array (MWA,
\citealt{tin13}), the Precision Array for Probing the Epoch of Reionization
(PAPER, \citealt{par10}), the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}), the LOw Frequency ARray (LOFAR,
\citealt{dev09}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME,
\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and the low
frequency component of the Square Kilometer Array Low Frequency Aperture Array
(SKA-Low \citealt{mel13}).

@@

There are some important differences between the two techniques:
\begin{enumerate}
\item The time-averaging cannot be performed on a stochastic measurement but
  only on its statistical properties. In FX imaging, the visibilities measured
  between antenna pairs represent spatial correlations which can be time-averaged
  followed by gridding and imaging. However, in MOFF imaging both
  antenna and gridded electric fields are stochastic and therefore must be
  imaged and squared before time-averaging. 
\item In FX imaging, electric fields measured by antennas are not correlated with
  themselves and hence lack zero spacing measurements. In contrast, in MOFF
  imaging, since the gridded electric fields are imaged and squared, they
  contain information from auto-correlated electric fields at zero spacing.
  Hence, they must be subtracted from the images.
\end{enumerate}

@@

\subsection{Calibration}
Calibration of direct imaging correlators remains an unresolved matter. Contrary
to the FX data flow, direct imagers mix the signals from all antennas before
averaging and writing to disk. It is therefore essential to apply gain solutions 
before the gridding step. Previous efforts have resorted to applying FX-generated
calibration solutions \citep{zhe14, fos14}, or integrating a dedicated FX correlator 
which periodically forms the full visibility matrix \citep{wij09, dev09}.

@@

In contrast, visibilities measured by an antenna are statistically stable within
an integration time interval. Hence, they are averaged after calibration as shown
in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities
before imaging because visibilities represent a compact representation of the
information in images. Hence it is more computationally efficient to average
antenna pair visibilities rather than images. Since this averaging has been
performed already on the visibilities over an integration timescale, the imaging
step has to be performed only once per integration cycle. FX imaging holds this
advantage as long as the square of the number of antennas is smaller than the
number of pixels on the image.

@@

The squaring operation under MOFF imaging in the image plane introduces antenna 
auto-correlations around the zero spacing in the $uv$-plane which are absent in 
traditional visibility-based imaging. In order to achieve a robust comparison, these 
auto-correlations have to be removed from both the $uv$ and image planes.

The shape and extent of these auto-correlations can be estimated from the 
antenna aperture illumination pattern which is a quantity already available 
from the gridding step. Fig.~\ref{fig:autocorr_wts_PB} shows the estimated 
weights from antenna auto-correlations in the $uv$-plane (left) and the 
corresponding response in the image plane (right). The latter is simply the 
directional antenna power response.

@@

@@

\subsection{Differences in outputs}\label{sec:diff}

@@

These residuals can be mitigated by: 
\begin{enumerate}
  \item making the grid spacing finer which makes the rounding error less 
    susceptible to the location of the antenna center relative to the grid, 
    and
  \item subtracting each auto-correlation of antenna weights separately by 
    using the shape and extent of the footprint appropriate for that specific
    antenna aperture. 
\end{enumerate}
The latter is the most general solution applicable especially in the case of 
heterogeneous antenna arrays.

The other notable difference is that weights outside the zero-spacing void in
some regions are different from each other at the few percent level though the 
sum of weights in these regions are identical. For instance, note the 
difference in weights at $(u,v)\approx (2,-10)$. This is found to arise 
because with MOFF imaging the antenna locations were rounded to the nearest 
grid pixel whereas in visibility-based imaging the baseline locations 
(difference between antenna locations) were rounded to the nearest grid pixel. 
Since rounding is not a commutative operation (i.e. rounding of the difference 
value is not necessarily equal to the difference of the rounded values), the 
gridding operation in the two cases introduces rounding errors in the placement 
of antenna aperture and cross-correlated aperture weights. Each antenna 
aperture weight projected to the nearest grid location can be displaced by 
$\sim 1$--2 pixels. This effect can be mitigated only by making the grid 
spacing finer.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.png}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using antenna-based MOFF algorithm (left) and FX 
    visibility-based software holography (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matches in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-uvwts}
\end{figure}

@@

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function uncorrected for the antenna power 
pattern. The rms of differenced synthesized beams is 
also modulated by the antenna power pattern. The rms of the difference is 
definitely lesser than the rms of the synthesized beam in the central regions 
up to $(l^2+m^2)^{1/2}\lesssim 0.3$. This implies that the beams are well 
matched in the central regions. In the outer regions, their mismatch is 
comparable to the rms of synthesized beams. This indicates the two 
synthesized beams are not completely randomly different from each other in
which case the rms of the difference would have been $\approx \sqrt{2}$
higher than the rms of the each of the synthesized beams. This indicates 
that while differences exist, large fractions of them are still well matched
to each other.Thus the rounding errors in gridding do not affect the 
statistics of the images or the synthesized beams.

@@

We have analyzed narrow band data from the LWA in New Mexico. This data is  
in LWA TBN format with 512 voltage time samples from 255 antennas within 
roughly a diameter of 100~m. The data is centered at a frequency of 74.03~MHz,
with a sample rate (equal to the bandwidth) of 100~kHz with 512 complex time 
samples in a A/D writeout timescale of 5.12~ms, a frequency resolution of 
195.3125~Hz and dual polarization. There are 391 such timestamps yielding a
total duration of 2~s.

We have matched the cable delays of all the antennas but yet to deploy the 
calibration pipeline we have developed on this data. Therefore, in 
subsequent discussion we assumed the data is calibrated.

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over the entire 2~s of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along x-axis and $m$ along y-axis. Due to lack of
calibration the flux scale is arbitrary. Despite the lack of calibration,
only with delay compensation, we see Cyg~A and Cas~A prominently at
$(l,m)\approx (0, 0.125)$, and (0.4,0.55) respectively and Sgr~A faintly at
(-0.5,-0.75), thus validating the functionality of the EPIC package.

\begin{figure}
  \includegraphics[width=\columnwidth]{LWA_sample_image_391_iterations.png}
  \caption{Image from LWA TBN data obtained with MOFF imaging using EPIC 
    package after averaging over 2~s and $\approx 80$~kHz. The x- and y-axes
    denote direction cosines $l$ and $m$ respectively. The antenna voltages
    are compensated for their respective delays. The flux scale is arbitrary
    and not calibrated. Cyg~A is prominently visible near the center at
    $(l,m)\approx (0, 0.125)$, Cas~A near (0.4,0.55) and Sgr~A near 
    (-0.5,-0.75).}
  \label{fig:LWA-image}
\end{figure}

@@

We have profiled the core routines of EPIC line-by-line for various ranges of
parameters such as antenna filling fraction, maximum baseline 
length, bandwidth and frequency resolution, integration timescale, etc. for 
HERA antenna layouts which are highly compact.

Of all steps in the MOFF pipeline that are repeated for every writeout from 
the F-engine, the slowest step even for dense HERA 
layouts is found to be the spatial two-dimensional FFT in the imaging stage 
relative to applying the sparse matrix gridding convolution, squaring or 
time-averaging. For instance, even in the dense array layout scenario that 
makes these stages perform the slowest, the gridding convolution, squaring and 
time-averaging take up only $\simeq 40$\%, $\simeq 40$\% and $\simeq 15$\% 
respectively of the time taken by the spatial Fourier transform. With sparser 
arrays they will be even faster.

@@

\begin{table}
  \centering
  \caption{Radio telescopes and array layouts.}
  \label{tab:antenna-layouts}
  \begin{threeparttable}
  \begin{tabular}{ccccc} 
    \hline
    Telescope & Core & Number & Antenna & Frequency \\
              & size & of Antennas & size & \\
    (key)     & $b_\textrm{max}$ (in m) & $N_\textrm{a}$ & $A_\textrm{a}$ (in m$^2$) & $f_0$ (in MHz) \\
    \hline
    MWA & 1400 & 112 & 16 & 150 \\
    (MWA-112) \\ \\
    MWA & 1400 & 496 & 16 & 150 \\
    (MWA-496) \\ \\
    LOFAR & 3500 & 24 & & 50 \\
    (LOFAR-C) & & & & \\ \\
    LWA1 & 100 & 256 & 10 & 50 \\
    (LWA1) \\ \\
    LWA-OV\tnote{a} & 200 & 256 & 10 & 50 \\
    (LWA-OV) & & & & \\ \\
    HERA & 70 & 19 & 154 & 150 \\
    (HERA-19) \\ \\
    HERA & 98 & 37 & 154 & 150 \\
    (HERA-37) \\ \\
    HERA & 294 & 331 & 154 & 150 \\
    (HERA-331) \\ \\
    HERA & 1330 & 6769\tnote{b} & 154 & 150 \\
    (HERA-6769) \\ \\
    SKA1-Low & 1000 & 750\tnote{c} & 962 & 150 \\
    (SKA1-LC) \\ \\
    SKA1-Low & 1000 & 192,000\tnote{c} & 962 & 150 \\
    Dipoles\tnote{d} \\
    (SKA1-LCD) \\
    \hline
  \end{tabular}
  \begin{tablenotes}
    \item[a] Owens Valley LWA
    \item[b] Hypothetically chosen to have a total collecting area of 
      1~km$^2$
    \item[c] This is the number of elements expected to be in the core, 
      three-fourths of the total number
    \item[d] All dipoles are used as independent elements without station 
      beamforming
  \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

Fig.~\ref{fig:parameter-space-bll-nant-instruments} shows the ratio of the
number of computations required with visibility based imaging relative to 
MOFF imaging and the boundaries where this ratio is unity. The color scale 
refers to the ratio while using a 1~m antenna element. Pairs of black and
white lines of the same line style correspond to a specific antenna size. For 
a given antenna size (annotated on the plot), the white line denotes the 
minimum baseline length that could achieve the closest packing as a function 
of the number of antennas. Regions to the right of the white line for a 
given antenna size imply a physically impossible scenario where antennas will 
have to be packed overlapping with each other. For the same antenna size, the 
black line denotes the boundary where the ratio of the number of computations 
with either algorithm becomes unity. Regions to the right of the black lines 
favour usage of the MOFF algorithm. Thus the wedge between the white and black 
lines denotes the region in parameter space where MOFF algorithm holds a 
definite advantage while visibility based algorithms will be favoured to the 
left of the black lines. As antenna size increases the maximum number of 
antennas for a dense packing as a function of baseline length decreases. Hence 
the white lines shift leftward as antenna size increases. Similarly, with 
increase in antenna size, $N_\textrm{g}$ also decreases when field of view 
imaging is achieved with an increasing grid spacing equal to antenna size and
hence lowers the amount of computations required with the MOFF algorithm. This
shifts the black curves leftward. HERA-331, HERA-1027, LWA1, SKA1-LC and 
SKA1-LCD are inside their respective wedges indicating they are in regions that
favour the usage of MOFF imaging. For fixed baseline length, regions favouring
the MOFF algorithm tend to be towards large $N_\textrm{a}$ indicating large-N
dense array layouts with smaller antenna elements are best suited for deploying 
the MOFF algorithm.

@@

% \subsection{Figures and tables}

% Figures and tables should be placed at logical positions in the text. Don't
% worry about the exact layout, which will be handled by the publishers.

% Figures are referred to as e.g. Fig.~\ref{fig:example_figure}, and tables as
% e.g. Table~\ref{tab:example_table}.

% % Example figure
% \begin{figure}
% 	% To include a figure from a file named example.*
% 	% Allowable file formats are eps or ps if compiling using latex
% 	% or pdf, png, jpg if compiling using pdflatex
% 	\includegraphics[width=\columnwidth]{example}
%     \caption{This is an example figure. Captions appear below each figure.
% 	Give enough detail for the reader to understand what they're looking at,
% 	but leave detailed discussion to the main body of the text.}
%     \label{fig:example_figure}
% \end{figure}

% % Example table
% \begin{table}
% 	\centering
% 	\caption{This is an example table. Captions appear above each table.
% 	Remember to define the quantities, symbols and units used.}
% 	\label{tab:example_table}
% 	\begin{tabular}{lccr} % four columns, alignment for each
% 		\hline
% 		A & B & C & D\\
% 		\hline
% 		1 & 2 & 3 & 4\\
% 		2 & 4 & 6 & 8\\
% 		3 & 5 & 7 & 9\\
% 		\hline
% 	\end{tabular}
% \end{table}

@@

We thank Larry D'Addario, Gregg Hallinan and Harish Vedantham for their valuable 
inputs.

>>>>

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts 
consisting of a large number of antennas 
($N_\textrm{a}\gtrsim 1000$--$\gtrsim 10000$). Since the cost of traditional 
correlators scales as $\mathcal{O}(N_\textrm{a}^2)$, there will be a steep cost for 
realizing the full imaging potential of these powerful instruments. Through 
our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present 
the first software demonstration of a generalized direct imaging algorithm 
known as the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost to $\mathcal{O}(N_\textrm{a}\log N_\textrm{a})$ 
but can also image from irregularly arranged heterogeneous antenna arrays. 
EPIC is highly modular and parallelizable. It is implemented in object oriented 
Python and is publicly available. We have verified the images produced to be 
mathematically identical to those produced using traditional techniques. We have 
also validated our implementation on data observed with the Long Wavelength 
Array. Antenna systems with a dense filling factor consisting of a large number 
of antennas such as LWA, SKA, and HERA will gain a significant advantage by 
deploying EPIC. Inherent availability of calibrated time-domain images on 
timescales roughly equal to the writeout timescale of the digitizer will make 
it a prime candidate for transient searches of primarily Fast Radio Bursts (FRB) 
as well as slower planteary and exoplantenary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction. A number of new telescopes are under development around the world
based on this new paradigm, including the Murchison Widefield Array (MWA;
\citealt{tin13,bow13}), the Precision Array for Probing the Epoch of Reionization
(PAPER; \citealt{par10}), the Hydrogen Epoch of Reionization 
Array\footnote{http://reionization.org} (HERA), the LOw Frequency ARray (LOFAR;
\citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME,
\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and the low
frequency component of the Square Kilometer Array Low Frequency Aperture Array
(SKA-Low \citealt{mel13}).

@@

There are some important differences between the two techniques:
\begin{enumerate}
\item The time-averaging cannot be performed on a stochastic measurement but
  only on its statistical properties. In FX imaging, the visibilities measured
  between antenna pairs represent spatial correlations which can be time-averaged
  followed by gridding and imaging. However, in MOFF imaging both antenna and 
  gridded electric fields are stochastic and therefore must be imaged and squared 
  before time-averaging. 
\item In FX imaging, electric fields measured by antennas are not correlated with
  themselves and hence lack zero spacing measurements. In contrast, in MOFF
  imaging, since the gridded electric fields are imaged and squared, they retain 
  information from auto-correlated electric fields at zero spacing and thus yield 
  the true total power of the imaged field.
\end{enumerate}

@@

\subsection{Calibration}
Calibration of direct imaging correlators remains a challenge. Contrary
to the FX data flow, direct imagers mix the signals from all antennas before
averaging and writing to disk. It is therefore essential to apply gain solutions 
before the gridding step. Previous efforts have resorted to applying FX-generated
calibration solutions \citep{zhe14,fos14}, or integrating a dedicated FX 
correlator which periodically forms the full visibility matrix 
\citep{wij09,dev09}.

@@

In contrast, visibilities measured by an antenna are statistically stable within
an integration time interval. Hence, they are averaged after calibration as shown
in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities
before imaging because the repeated cost of spatial FFT can be avoided. Since 
this averaging has been performed already on the visibilities over an integration 
timescale, the imaging step has to be performed only once per integration cycle. 
% FX imaging holds this advantage as long as the square of the number of antennas 
% is smaller than the number of pixels on the image.

@@

The squaring operation under MOFF imaging in the image plane introduces antenna 
auto-correlations around the zero spacing in the $uv$-plane which are absent in 
traditional visibility-based imaging. In order to facilitate a robust comparison, 
these auto-correlations are removed from both the $uv$ and image planes, which 
is otherwise not an essential part of the core algorithm. We describe below how 
they can be removed from the $uv$ plane in a straightforward manner.

The shape and extent of these auto-correlations can be estimated from the 
antenna aperture illumination pattern. The aperture illumination patterns
are already available from the gridding step. Fig.~\ref{fig:autocorr_wts_PB} 
shows the estimated weights from antenna auto-correlations in the $uv$-plane 
(left) and the corresponding response in the image plane (right). The latter 
is simply the directional antenna power response.

@@

In order to prove the accuracy of the EPIC code, we verify the images produced 
through simulations. We simulate electric field streams from a model sky and 
process the data through both the MOFF and visibility based imaging algorithms. 
We then compare the output images to demonstrate their equivalence.

@@

\subsection{Comparison of outputs}\label{sec:diff}

@@

These residuals can be mitigated by: 
\begin{enumerate}
  \item making the grid spacing finer which makes the rounding error less 
    susceptible to the location of the antenna center relative to the grid, 
    and
  \item subtracting each auto-correlation of antenna weights separately by 
    using the shape and extent of the footprint appropriate for that specific
    antenna aperture. 
\end{enumerate}
The latter is the most general solution applicable especially in the case of 
heterogeneous antenna arrays and is under active development for EPIC.

The other notable difference is that weights outside the zero-spacing void in
some regions are different from each other at the few percent level though the 
sum of weights in these regions are identical. For instance, note the 
difference in weights at $(u,v)\approx (2,-10)$. This is found to arise 
because with MOFF imaging the antenna locations were rounded to the nearest 
grid pixel whereas in visibility-based imaging the baseline locations 
(difference between antenna locations) were rounded to the nearest grid pixel. 
Since rounding is not a commutative operation (i.e. rounding of the difference 
value is not necessarily equal to the difference of the rounded values), the 
gridding operation in the two cases introduces rounding errors in the placement 
of antenna aperture and cross-correlated aperture weights. Each antenna 
aperture weight projected to the nearest grid location can be displaced by 
$\sim 1$--2 pixels. This effect can be mitigated only by making the grid 
spacing finer at the expense of increased computational cost.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.png}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using antenna-based MOFF algorithm (left) and FX 
    visibility-based software holography (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matched in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-uvwts}
\end{figure}

@@

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function uncorrected for the antenna power 
pattern. The rms of differenced synthesized beams is 
also modulated by the antenna power pattern. The rms of the difference is 
definitely lesser than the rms of the synthesized beam in the central regions 
up to $(l^2+m^2)^{1/2}\lesssim 0.3$. This implies that the beams are well 
matched in the central regions. In the outer regions, their mismatch is 
comparable to the rms of synthesized beams. This indicates the two 
synthesized beams are not completely randomly different from each other in
which case the rms of the difference would have been $\approx \sqrt{2}$
higher than the rms of the each of the synthesized beams. This indicates 
that while differences exist, large fractions of them are still well matched
to each other even out to the horizon. Thus the rounding errors in gridding 
do not affect the statistics of the images or the synthesized beams.

@@

Here we demonstrate our software using narrow band data from the LWA in New 
Mexico. This data is in LWA narrow-band transient buffer (TBN) format with 512 
voltage time samples from 255 antennas within roughly a diameter of 100~m. The 
data is centered at a frequency of 74.03~MHz, with a sample rate (equal to the 
bandwidth) of 100~kHz with 512 complex time samples in a A/D writeout timescale 
of 5.12~ms, a frequency resolution of 195.3125~Hz and dual polarization. There 
are 391 such timestamps yielding a total duration of 2~s.

We corrected the cable delays, but otherwise assume the data is sufficiently 
calibrated to image directly. A test of EPICal on this data will be conducted 
in the future.

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over the entire 2~s of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along x-axis and $m$ along y-axis. The flux scale 
is arbitrary. Even in this proof-of-concept demonstration, we see Cyg~A and 
Cas~A prominently and Sgr~A faintly as annotated, thus validating the 
functionality of the EPIC package.

\begin{figure}
  \includegraphics[width=\columnwidth]{LWA_MOFF_bandavg_image_4_iterations_analytic_aperture.eps}
  \caption{Image from LWA TBN data obtained with MOFF imaging using EPIC 
    package after averaging over 2~s and $\approx 80$~kHz. The x- and y-axes
    denote direction cosines $l$ and $m$ respectively. The antenna voltages
    are compensated for their respective delays. The flux scale is arbitrary
    and not calibrated. Locations of Cyg~A, Cas~A, Sgr~A, Her~A and Jupiter
    are annotated.}
  \label{fig:LWA-image}
\end{figure}

@@

We have profiled the core routines of EPIC line-by-line for various ranges of 
parameters such as antenna filling fraction, maximum baseline length, bandwidth 
and frequency resolution, integration timescale, etc. for HERA antenna layouts 
which are highly compact. However, we note that in general, the hardware and 
optimization of routines in place will determine the relative speeds of the 
different stages in the pipeline.

Of all steps in the MOFF pipeline that are repeated for every writeout from 
the F-engine, the slowest step even for dense HERA layouts is found to be the 
spatial two-dimensional FFT in the imaging stage relative to applying the sparse 
matrix gridding convolution, squaring or time-averaging. For instance, even in 
the dense array layout scenario that makes these stages perform the slowest, the 
gridding convolution, squaring and time-averaging take up only $\simeq 40$\%, 
$\simeq 40$\% and $\simeq 15$\% respectively of the time taken by the spatial 
Fourier transform. With sparser arrays the gridding process will be be even 
faster.

@@

\begin{table}
  \centering
  \caption{Radio telescopes and array layouts.}
  \label{tab:antenna-layouts}
  \begin{threeparttable}
  \begin{tabular}{ccccc} 
    \hline
    Telescope & Core & Number & Antenna & Frequency \\
              & size & of Antennas & size & \\
              & $b_\textrm{max}$ (in m) & $N_\textrm{a}$ & $A_\textrm{a}$ (in m$^2$) & $f_0$ (in MHz) \\
    \hline
    MWA-112\tnote{a} & 1400 & 112 & 16 & 150 \\
    MWA-240\tnote{a} & 1400 & 240 & 16 & 150 \\
    MWA-496\tnote{a} & 1400 & 496 & 16 & 150 \\
    MWA-112\tnote{a} & 1400 & 1008 & 16 & 150 \\
    LOFAR-LC\tnote{b} & 3500 & 24 & 5809 & 50 \\
    LOFAR-HC\tnote{b} & 3500 & 48 & 745 & 150 \\
    LWA1 & 100 & 256 & 10 & 50 \\
    LWA-OV\tnote{c} & 200 & 256 & 10 & 50 \\
    HERA-19 & 70 & 19 & 154 & 150 \\
    HERA-37 & 98 & 37 & 154 & 150 \\
    HERA-331 & 294 & 331 & 154 & 150 \\
    HERA-6769\tnote{d} & 1330 & 6769 & 154 & 150 \\
    SKA1-LC\tnote{e} & 1000 & 750 & 962 & 150 \\
    SKA1-LCD\tnote{f} & 1000 & 192,000 & 962 & 150 \\
    CHIME & 100 & 1280 & 8 & 600 \\
    HIRAX & 200 & 1024 & 6 & 600 \\
    \hline
  \end{tabular}
  \begin{tablenotes}
    \item[a] MWA-N denotes N tiles in the specified core diameter
    \item[b] LC and HC denotes low band and high band stations inside the 
      specified core diameter 
    \item[c] Owens Valley LWA
    \item[d] Hypothetically chosen to have a total collecting area of 
      1~km$^2$
    \item[e] This is the number of beamformed stations expected to be in the 
      core, roughly three-fourths of the total number
    \item[f] All dipoles inside the core are used as independent elements 
      without station beamforming
  \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

Fig.~\ref{fig:parameter-space-bll-nant-instruments} shows the ratio of the
number of computations required with visibility based imaging relative to 
MOFF imaging and the boundaries where this ratio is unity. The color scale 
refers to the ratio while using a 1~m antenna element. Pairs of black and
white lines of the same line style correspond to a specific antenna size. For 
a given antenna size (annotated on the plot), the white line denotes the 
minimum baseline length that could achieve the closest packing as a function 
of the number of antennas. Regions to the right of the white line for a 
given antenna size imply a physically impossible scenario where antennas will 
have to be packed overlapping with each other. For the same antenna size, the 
black line denotes the boundary where the ratio of the number of computations 
with either algorithm becomes unity. Regions to the right of the black lines 
favour usage of the MOFF algorithm. Thus the wedge between the white and black 
lines denotes the region in parameter space where MOFF algorithm holds a 
definite advantage while visibility based algorithms will be favoured to the 
left of the black lines. As antenna size increases the maximum number of 
antennas for a dense packing as a function of baseline length decreases. Hence 
the white lines shift leftward as antenna size increases. Similarly, with 
increase in antenna size, $N_\textrm{g}$ also decreases when field of view 
imaging is achieved with an increasing grid spacing equal to antenna size and
hence lowers the amount of computations required with the MOFF algorithm. This
shifts the black curves leftward. HERA-331, HERA-1027, LWA1, SKA1-LC and 
SKA1-LCD are inside their respective wedges indicating they are in regions that
favour the usage of MOFF imaging. For fixed baseline length, regions favouring
the MOFF algorithm tend to be towards large $N_\textrm{a}$ indicating large-N
dense array layouts with smaller antenna elements are best suited for deploying 
EPIC.

@@

@@

We thank Greg Taylor for providing us the LWA data, and Larry D'Addario, 
Gregg Hallinan and Harish Vedantham for their valuable inputs. Construction of 
the LWA has been supported by the Office of Naval Research under Contract 
N00014-07-C-0147. Support for operations and continuing development of the LWA1 
is provided by the National Science Foundation under grant AST-1139974 of the 
University Radio Observatory program.
